LabMate: a prospectus for types for MATLAB

### 2024-04-12 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] section 4 expanded

\section{Implementation}

@@

LabMate elaborates MATLAB expressions and commands into terms of its own internal core type theory.
%
Because we want to ergonomically handle physical dimensions, matrices
and their types, this type theory has a richer equational theory than
other state-of-the-art type theories, which we achieve by a refined
normalisation algorithm for free Abelian groups, and free monoids.
%
The actual typechecking is implemented as a stack based virtual machine which uses unification and a limited form of backtracking to solve typing constraints.
%
After the virtual machine has solved as many problems as it can, we analyse its final state to reconstruct the output source code, perhaps with inserted warnings and error messages, or additional responses to queries.

>>>>

\section{Implementation and Current Status}

@@

The meaning of whitespace in MATLAB is highly context-sensitive --- e.g., $\mbox{\texttt{(2 - 2)}}$, $\mbox{\texttt{[2 - 2]}}$ and $\mbox{\texttt{(2 -2)}}$ all meaning zero, but $\mbox{\texttt{[2 -2]}}$ being a $2\times1$ matrix.
%
Another precursor to typechecking is the deduction of variable scope: as proof of concept, we implemented the directive
\begin{verbatim}
%> rename x y
\end{verbatim}
which requests LabMate to rename all and only the occurrences of the \texttt{x} in scope at that point to \texttt{y}, but only if that does not conflict with other things called \texttt{y}.
%
To our surprise, we discovered that a similar facility offered by MATLAB's own editor sometimes changes the semantics of programs, which is presumably not what the user intends.
%
Writing a typechecker, we cannot afford the luxury of waiting until runtime to figure out which $\texttt{x}$ is which.

LabMate elaborates MATLAB expressions and commands into terms of its own internal core type theory.
%
Although the simple examples in this paper use \emph{constant} sizes and dimensions, any practicable library requires size, header and dimension expressions to contain \emph{parameters}.
%
For example, if \texttt{A} is $i\times j$ and \texttt{B} is $i\times k$, then
\begin{verbatim}
[ A B
  B A ]
\end{verbatim}
should be a valid matrix pasting, because $j+k = k+j$. Our notion of equality for numbers thus induces commutative monoid structure.
%
Physical dimensions are modelled as the free Abelian group generated by some base set of dimensions~\cite{dim.group}, so we decide that theory, too.
%
The core theory we need is in place.

The actual typechecking is implemented as a stack based virtual machine which traverses the syntax tree, deducing its translation into the core theory by propagating type information, generating typing constraints, and solving them by unification~\cite{robinson65}.
%
In order to support a much more expressive language of types, we must move on from the classic requirement that all constraints can be solved or rejected on first inspection and hence that the program can be processed exactly once and in a specific order~\cite{damas.milner}.
%
Our virtual machine supports repeated refocusing to wherever in the program there is progress to be made.
%
Moreover, as MATLAB notation is somewhat overloaded (e.g., \texttt{*} can mean either scalar or matrix multiplication), so LabMate elaboration has a (limited) need for backtracking search, so that, too, is facilitated.

Elaboration for matrix multiplication and addition currently generate the correct constraints, but to check our examples, we need to improve our constraint solver and also implement the $\bsl$ operator.
%
As division is specified in terms of multiplication, its typing discipline requires a similar coincidence of header lists, but for the rows of both inputs: the technology we need is more of the same.

After the virtual machine has solved as many problems as it can, we analyse its final state to reconstruct the output source code, perhaps with inserted warnings and error messages, or additional responses to queries and generated code.
